---
title: "feature_exploration"
output: html_document
---

```{r}
library(dplyr)
library(ggplot2) ## graphs 
library(Amelia) ## missing data 
library(caret)
library(tidyverse)
library(ggplot2)
```

```{r}
df <- read.csv(file='../data/260_1123cohort.csv',sep=',', header=TRUE, quote="\'", check.names = FALSE)

# pre process the names of columns
colnames(df) <- gsub('\"', '', colnames(df))
colnames(df)[1] <- "index"

#preprocess the names of rows
rownames(df) <- gsub('"', '', df$index)
drops <- c("index")
df <- df[ , !(names(df) %in% drops)]

#head of dataframe
df %>% head()
```

Preprocess the values of some columns to get rid of the quotes:
```{r}
df$ICUSTAY_AGE_GROUP <- gsub('"', '', df$ICUSTAY_AGE_GROUP)
df$first_icu_stay <- gsub('"', '', df$first_icu_stay)
df$gender <- gsub('""', '', df$gender)
df$admission_type <- gsub('""', '', df$admission_type)
```


Visualization of missing data:
```{r}
missmap(df)
```

Values to check: temperature, urine

Get rid of patients that have missing gender:
```{r}
df <- df %>% filter(!is.na(gender))
```

We can start working with columns that have less than 20% of missing data:
```{r}
threshold <- nrow(df)*0.20
df2 <- df[colSums(is.na(df)) < threshold]
# summary(df2)
```

```{r}
missmap(df2)
```

Remove patients that have lots of missing values:
```{r}
# threshold <- length(df2)*0.0001
df3 <- df2[rowSums(is.na(df2)) < 1,]
# summary(df3)
missmap(df3)
```


```{r}
df4<-df3 %>% 
  select(-c("subject_id","hadm_id","ICUSTAY_AGE_GROUP","hospital_expire_flag","OASIS","OASIS_PROB"))

summary(df4)

```
```{r}
df4 %>% ggplot(aes(x=age,y=..density..))+
  geom_histogram()
```

Pre-processing of age:
-replace values bigger than 200 by 91.5
```{r}
df4 <- df4 %>% 
  mutate(age=ifelse(age>=90,91.5,age))

df4 %>% ggplot(aes(x=age,y=..density..))+
  geom_histogram(col="grey")
```

Pre-processing categorical variable into 0vs1 columns and making dummies:
```{r}
df5<-df4 %>% 
  mutate(gender=ifelse(gender=="F", 1,0),# 1 is female and 0 is male
         first_icu_stay=ifelse(first_icu_stay==FALSE,0,1),
         v=1, # needed to make dummies 
         admission_type=paste("AdmissionType",admission_type,sep="_")) %>% 
  mutate(first_icu_stay=ifelse(first_icu_stay==FALSE,0,1)) %>%
  spread(admission_type, v, fill = 0) %>%
  filter(age>17) %>% ## leave only adults 
  unique() %>% ## unique entries 
  mutate(urineoutput=abs(urineoutput)) %>% ## we have negative values. I change them to positive
  
```


Optional if you want to reduce outliers 
```{r}
df6<-df5 %>%  
  select(HeartRate_Min:RespRate_Mean) %>%
  gather(variable, value) %>%
  group_by(variable) %>%
  mutate(mean=mean(value), 
         twosd=(sd(value))*2,
         lower=mean-twosd,
         upper=mean+twosd) %>%
  select(-c(value))%>%
  unique()

df7<-df5 %>% 
  select(icustay_id,HeartRate_Min:RespRate_Mean) %>%
  gather(variable, value, -icustay_id)

df8<-left_join(df7,df6)%>%
  filter(value>=lower & value<=upper)%>%
  select(icustay_id, variable, value)%>%
  spread(variable, value)

df8 <- df8[rowSums(is.na(df8)) < 1,]
df8 <- df8 %>% select(icustay_id)%>% unique()

df9<- left_join(df8, df5)

summary(df5$icustay_expire_flag)
summary(df9$icustay_expire_flag)
```






Write the df5 file into a csv file:
```{r}
write.csv(df5, '../data/cohort_preprocessed.csv')
```















Visualisation Lab Results
```{r}
labresults<- df5 %>%
  select(icustay_id, icustay_expire_flag, ANIONGAP_min:WBC_max)%>%
  gather(., test, value, -icustay_expire_flag, -icustay_id)%>%
  separate(test, into = c("test", "MinOrMax"), sep="_")

plot<-labresults %>% 
  split(.$test) %>%
  map(~ggplot(.,aes(x=value, col=MinOrMax))+
  geom_density()+
    theme_bw()+
ggtitle(paste("distributions",names(.x%>%split(.$test)))))
paths <- stringr::str_c("LabtestDistribution",names(plot), ".jpg")
pwalk(list(paths, plot), ggsave, path = '../graphs')

labresults<- labresults %>% 
  spread(., MinOrMax, value)

plot2<-labresults %>% 
  mutate(died=ifelse(icustay_expire_flag==1, "Died", "Survived"))%>%
  group_by(min, max, test, died)%>%
  mutate(howmany=n())%>%
  ungroup()%>%
  split(.$test) %>%
  map(~ggplot(.,aes(x=max, y=min, size=howmany, col=died))+
  geom_point(alpha=0.2)+
    theme_bw()+
ggtitle(paste("Scatter plot for",names(.x%>%split(.$test)))))

plot2[1]
paths <- stringr::str_c("LabtestScatter",names(plot2), ".jpg")
pwalk(list(paths, plot2), ggsave, path = '../graphs')


plot3<-labresults %>% 
  mutate(died=ifelse(icustay_expire_flag==1, "Died", "Survived")) %>%
group_by(min, max, test,died)%>%
  mutate(howmany=n())%>%
  ungroup()%>%
  mutate(facet=died)%>%
  uncount(2, .id="id")%>%
  mutate(died=ifelse(id==2 & died=="Died", "Died*", died))%>%
  mutate(died=ifelse(id==2 & died=="Survived", "Survived*", died))%>%
  mutate(facet=ifelse(id==1, died, "both"))%>%
  split(.$test) %>%
  map(~ggplot(.,aes(x=max, y=min, size=howmany, colour=died))+
  geom_point(alpha=0.2)+
    scale_color_manual(breaks = c("Died", "Died*", "Survived", "Survived*"),
                        values=c("red", "red", "green", "green"))+
    theme_bw()+
    facet_grid(~facet)+
ggtitle(paste("Scatter plot faceted for",names(.x%>%split(.$test)))))

plot3[1]
paths <- stringr::str_c("LabtestScatterFacet",names(plot3), ".jpg")
pwalk(list(paths, plot3), ggsave, path = '../graphs')
```




Visualisation Vitals
```{r}
Vitals<- df5 %>%
  select(icustay_id, icustay_expire_flag, heartrate:temp_score,HeartRate_Min:RespRate_Mean) %>%
  mutate(icustay_expire_flag=as.character(icustay_expire_flag)) %>%
  gather(vital, value, -icustay_id, -icustay_expire_flag)


## jitters
Plots<-Vitals %>% 
  split(.$vital) %>%
  map(~ggplot(.,aes(x=icustay_expire_flag, y=value, alpha=0.1))+
        ggtitle(paste("Variable vs Outcome for ",names(.x%>%split(.$vital))))+
        geom_jitter(width = 0.3))
Plots
paths <- stringr::str_c("Jitter",names(Plots), ".jpg")
pwalk(list(paths, Plots), ggsave, path = '../graphs')




```


Visualisation Lab Results
```{r}
labresults<- df5 %>%
  select(icustay_id, icustay_expire_flag, ANIONGAP_min:WBC_max)%>%
  gather(., test, value, -icustay_expire_flag, -icustay_id)%>%
  separate(test, into = c("test", "MinOrMax"), sep="_")

plot<-labresults %>% 
  split(.$test) %>%
  map(~ggplot(.,aes(x=value, col=MinOrMax))+
  geom_density()+
    theme_bw()+
ggtitle(paste("distributions",names(.x%>%split(.$test)))))
paths <- stringr::str_c("LabtestDistribution",names(plot), ".jpg")
pwalk(list(paths, plot), ggsave, path = '../graphs')

labresults<- labresults %>% 
  spread(., MinOrMax, value)

plot2<-labresults %>% 
  mutate(died=ifelse(icustay_expire_flag==1, "Died", "Survived"))%>%
  group_by(min, max, test, died)%>%
  mutate(howmany=n())%>%
  ungroup()%>%
  split(.$test) %>%
  map(~ggplot(.,aes(x=max, y=min, size=howmany, col=died))+
  geom_point(alpha=0.2)+
    theme_bw()+
ggtitle(paste("Scatter plot for",names(.x%>%split(.$test)))))

plot2[1]
paths <- stringr::str_c("LabtestScatter",names(plot2), ".jpg")
pwalk(list(paths, plot2), ggsave, path = '../graphs')


plot3<-labresults %>% 
  mutate(died=ifelse(icustay_expire_flag==1, "Died", "Survived")) %>%
group_by(min, max, test,died)%>%
  mutate(howmany=n())%>%
  ungroup()%>%
  mutate(facet=died)%>%
  uncount(2, .id="id")%>%
  mutate(died=ifelse(id==2 & died=="Died", "Died*", died))%>%
  mutate(died=ifelse(id==2 & died=="Survived", "Survived*", died))%>%
  mutate(facet=ifelse(id==1, died, "both"))%>%
  split(.$test) %>%
  map(~ggplot(.,aes(x=max, y=min, size=howmany, colour=died))+
  geom_point(alpha=0.2)+
    scale_color_manual(breaks = c("Died", "Died*", "Survived", "Survived*"),
                        values=c("red", "red", "green", "green"))+
    theme_bw()+
    facet_grid(~facet)+
ggtitle(paste("Scatter plot faceted for",names(.x%>%split(.$test)))))

plot3[1]
paths <- stringr::str_c("LabtestScatterFacet",names(plot3), ".jpg")
pwalk(list(paths, plot3), ggsave, path = '../graphs')



